---
title: "Predict_Quant_from_NPX_Vignettes"
author: "Jingjing Qi, Icahn School of Medicine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predict_Quant_from_NPX_Vignettes}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=F}
library(tidyverse) 
library(NPXtools)
#library(magrittr)
#library(tidyr)
#library(SummarizedExperiment)
```

## generat fitting curve 
use a list object of npx and a list object of quant paired with the same <name>.    
at least 1 file is required in the list, at least 1 analyte/plate has both npx and quant value.     

this workbook use run001-run010 empirical data generated the standard curve.    
predicted quant value and the targeted value was compared use run011-020    

```{r, eval=F}
my_Bucket <- bucketlist()
#---
my_Bucket <- my_Bucket[my_Bucket$Bucket == "himc-project-data", 1]
my_Bucket_file <- get_bucket(my_Bucket, prefix = "MSC", max = Inf)
f.list <- sapply(my_Bucket_file, function(x){
  x[[1]]
}) 
f.list <- f.list[grepl(".*NPX.xlsx", f.list)]
names(f.list) <- f.list

# QUANT file format: startrow = 9
# NPX file format: startrow = 8
f.list <- lapply(f.list, function(x){
  temp <- try(s3read_using(FUN = read_npx, object = x, bucket = my_Bucket, startrow = 8))
  temp$f_name <- x
  temp
})


f.list.quant <- sapply(my_Bucket_file, function(x){
  x[[1]]
}) 
f.list.quant <- f.list.quant[grepl(".*QUANT.xlsx", f.list.quant)]
names(f.list.quant) <- f.list.quant

# QUANT file format: startrow = 9
# NPX file format: startrow = 8
f.list.quant <- lapply(f.list.quant, function(x){
  temp <- try(s3read_using(FUN = read_npx, object = x, bucket = my_Bucket, startrow = 9, type = "QUANT"))
  temp$f_name <- x
  temp
})

names(f.list) <- gsub("(.*/|_QUANT.xlsx|_NPX.xlsx)", "", names(f.list) )
names(f.list.quant) <- gsub("(.*/|_QUANT.xlsx|_NPX.xlsx)", "", names(f.list.quant) )
names(f.list) == names(f.list.quant)


make.names(rownames(f.list[[1]]))
marker <- "IL17F"
quant_npx <- lapply(make.names(rownames(f.list[[1]])), function(y){
  lapply(names(f.list), function(x){
  data.frame(NPX = f.list[[x]]@assays@data$npx[rownames(f.list[[x]]) == y, ],
             Quant = f.list.quant[[x]]@assays@data$npx[rownames(f.list.quant[[x]]) == y, ],
             run = x,
             lod = f.list[[x]]@elementMetadata$LOD[rownames(f.list.quant[[x]]) == y]%>%as.numeric())
   })%>%
     do.call(what = rbind)%>%
     mutate(analyt = y)
})%>%
  do.call(what = rbind)

lod <- distinct(quant_npx)%>%
  select(analyt, lod)%>%
  group_by(analyt)%>%
  summarise(lod = median(lod, na.rm = T))

# npx to quant association plot  is.na(quant) is in the range of avaliable npx
quant_npx%>%
  mutate(Quant = ifelse(is.na(Quant), 0, Quant))%>%
  ggplot()+
  geom_point(aes(NPX, log10(Quant), color = run), shape = 21)+
  geom_vline(aes(xintercept = lod), data = lod, color = "red")+
  facet_wrap(~analyt, ncol = 10, scales = "free")+
  theme_bw()


lapply(names(f.list), function(x){
  try(lm(log10(Quant) ~ NPX, na.action = na.exclude,
     data = quant_npx%>%
       filter(run == x)))
})

analyts <- unique(quant_npx$analyt)
names(analyts) <- analyts

# distribution of NPX
#cor_plots <- lapply(analyts, function(x){
#  data.frame(NPX = quant_npx$NPX,
#             Quant = quant_npx$Quant,
#             analyts = quant_npx$analyt,
#             run = quant_npx$run)%>%
#    filter(run == "MSC_Target48_Run001", analyts == x)%>%
#    filter(NPX > quantile(NPX, probs = 0.02, na.rm = T), NPX < quantile(NPX, probs = 0.98, na.rm = T))%>%
#    ggplot()+
#    #geom_density(aes(x = NPX))+
#    #geom_density(aes(x = Quant))+
#    geom_density(aes(x = log10(Quant)))+
#    labs(title = x)+
#    theme_bw()
#})
#wrap_plots(cor_plots, ncol = 10)
#

fit_list_48plx <- lapply(analyts, function(x){
  df <- quant_npx%>%
    filter(analyt == x, !is.na(Quant), !is.na(NPX))
  if(x == "IL4"){
    lm(log10(Quant) ~ splines::ns(NPX, knots = c(1, 2, 4)), data = df) # smoothspine didnt converge changed to fix knots spline, validated with test data.
  }else{
    smooth.spline(y = log10(df$Quant), x = df$NPX, cv = T)
  }
    
}) 


#saveRDS(fit_list_48plx, file = "~/Downloads/fit_list_48plx_run001_to_run010.rds")
```

## pipeline predict quant using npx
use a list object of npx and a list object of quant paired with the same <name>.   
According to Olink, the STD curve binds to a given kit lot.     

```{r, eval=F}
library(aws.s3)

Sys.setenv()

my_Bucket <- bucketlist()
#---
my_Bucket <- my_Bucket[my_Bucket$Bucket == "himc-project-data", 1]
my_Bucket_file <- get_bucket(my_Bucket, prefix = "MSC", max = Inf)

f.list <- sapply(my_Bucket_file, function(x){
  x[[1]]
}) 
f.list <- f.list[grepl(".*NPX.xlsx", f.list)]
names(f.list) <- f.list

# QUANT file format: startrow = 9
# NPX file format: startrow = 8
f.list <- lapply(f.list, function(x){
  temp <- try(s3read_using(FUN = read_npx, object = x, bucket = my_Bucket, startrow = 8))
  temp$f_name <- x
  temp
})


f.list.quant_to_replace <- sapply(my_Bucket_file, function(x){
  x[[1]]
}) 
f.list.quant_to_replace <- f.list.quant_to_replace[grepl(".*QUANT.xlsx", f.list.quant_to_replace)]
names(f.list.quant_to_replace) <- f.list.quant_to_replace

# QUANT file format: startrow = 9
# NPX file format: startrow = 8
f.list.quant_to_replace <- lapply(f.list.quant_to_replace, function(x){
  temp <- try(s3read_using(FUN = read_xlsx, object = x, bucket = my_Bucket, sheet = 1, col_names = F))
  temp
})

names(f.list) <- gsub("(.*/|_QUANT.xlsx|_NPX.xlsx)", "", names(f.list) )
names(f.list.quant_to_replace) <- gsub("(.*/|_QUANT.xlsx|_NPX.xlsx)", "", names(f.list.quant_to_replace) )
names(f.list) == names(f.list.quant_to_replace)

fit_list_48plx <- readRDS("~/Downloads/fit_list_48plx_run001_to_run010.rds")
```

## generate predicted file     
output:    
save Quant csv with name: list<name>__Quant_Predicted.csv    


```{r, eval=F}
names(f.list)
#plt2pdt <- "MSC_Target48_Run009"
pre_fix <- "~/Downloads/"

pdt_quant <- sapply(rownames(f.list[[plt2pdt]]), function(x){
  test <- f.list[[plt2pdt]]@assays@data$npx[rownames(f.list[[plt2pdt]]) == x, ]
  na_idx <- which(is.na(test))
  test[na_idx] <- 1
  if(x == "IL4"){
    re <- predict(fit_list_48plx[[x]], newdata = list(NPX = test))
  }else{
    re <- predict(fit_list_48plx[[x]], x = test)$y
  }
  re[na_idx] <- NA
  round(10^re, 5) # back to origional scale
})

quant_to_replace <- f.list.quant_to_replace[[plt2pdt]]
npx_mat <- t(f.list[[plt2pdt]]@assays@data$npx)

for (x in colnames(pdt_quant)) {
  idx <- which(quant_to_replace[4, ] == x)
  col_to_replace <- quant_to_replace[9 : (9 + nrow(pdt_quant) - 1), idx]
  idx_char <- which(is.na(sapply(col_to_replace, as.numeric)))
  
  replace_col <- as.character(pdt_quant[, x])
  
  npx_check <- npx_mat[ , x]
  
  if(sum(is.na(npx_check)) == nrow(col_to_replace)){
    replace_col <- "No Data"
  }
  if(!is.null(idx_char) & (length(idx_char) != nrow(col_to_replace))){
    for (i in idx_char) {
      replace_col[i] <- unlist(col_to_replace)[i]
    }
  }else{
    quant_to_replace[9 : (9 + nrow(pdt_quant) - 1), idx] <- replace_col
  }
}

colnames(quant_to_replace) <- NULL
f_name <- paste0(pre_fix, plt2pdt, "_Quant_Predicted.csv")
write.csv(quant_to_replace, f_name, row.names = F, na = "")
```

## validate correlation of predicted quant and software exported quant

```{r, eval=F}
test = read.xlsx("~/Downloads/pngparsed_analyt(AutoRecovered).xlsx", sheetIndex = 1)
colnames(test)
plt2pdt <- "MSC_Target48_Run012"
r2_pdt_quant <- sapply(analyts, function(x){
  if(x == "IL4"){
    predict(fit_list_48plx[[x]], newdata = list(NPX = f.list[[plt2pdt]]@assays@data$npx[rownames(f.list[[plt2pdt]]) == x, ]))
  }else{
    predict(fit_list_48plx[[x]], x = f.list[[plt2pdt]]@assays@data$npx[rownames(f.list[[plt2pdt]]) == x, ])$y
  }
})

f.list[[plt2pdt]]$Assay
dim(f.list[[plt2pdt]])


cor_plots <- lapply(analyts, function(x){
  data.frame(Quant_software = log10(f.list.quant[[plt2pdt]]@assays@data$npx[rownames(f.list[[plt2pdt]]) == x, ]),
             Quant_predicted = r2_pdt_quant[ , colnames(r2_pdt_quant) == x])%>%
  ggscatter(x = "Quant_software", y = "Quant_predicted", palette = "jco",
   add = "reg.line", conf.int = F)+
   stat_cor(label.x.npc = 0.2, color = "red")+
    labs(title = x)+
    theme_bw()
})

wrap_plots(cor_plots, ncol = 10)

marker <- "IL10"
data.frame(Quant_screenshot = test$Run002_IL10_Quanti[-c(24, 30, 36, 42, 48)],
             Quant_predicted = 10^r2_pdt_quant[ , colnames(pdt_quant) == marker])%>%
  ggscatter(x = "Quant_screenshot", y = "Quant_predicted", palette = "jco",
   add = "reg.line", conf.int = F)+
   stat_cor(label.x.npc = 0.2, color = "red")+
    labs(title = "Run002_IL10_Quanti")+
    theme_bw()

```

## validate correlation of predicted quant and software screenshot quant

```{r, eval=F}
colnames(test)
plt2pdt <- "MSC_Target48_Run006"
r2_pdt_quant <- sapply(analyts, function(x){
  if(x == "IL4"){
    predict(fit_list_48plx[[x]], newdata = list(NPX = f.list[[plt2pdt]]@assays@data$npx[rownames(f.list[[plt2pdt]]) == x, ]))
  }else{
    predict(fit_list_48plx[[x]], x = f.list[[plt2pdt]]@assays@data$npx[rownames(f.list[[plt2pdt]]) == x, ])$y
  }
})

f.list[[plt2pdt]]$Assay
dim(f.list[[plt2pdt]])

marker <- "IL10"
data.frame(Quant_screenshot = test$Run006_IL10_Quanti[-c(24, 30, 36, 42, 48)][-6],
             Quant_predicted = 10^r2_pdt_quant[ , colnames(pdt_quant) == marker])%>%
  ggscatter(x = "Quant_screenshot", y = "Quant_predicted", palette = "jco",
   add = "reg.line", conf.int = F)+
   stat_cor(label.x.npc = 0.2, color = "red")+
    labs(title = "Run006_IL10_Quanti")+
    theme_bw()

marker <- "IL2"
data.frame(Quant_screenshot = test$Run006_IL2_Quanti[-c(24, 30, 36, 42, 48)][-6],
             Quant_predicted = 10^r2_pdt_quant[ , colnames(pdt_quant) == marker])%>%
  ggscatter(x = "Quant_screenshot", y = "Quant_predicted", palette = "jco",
   add = "reg.line", conf.int = F)+
   stat_cor(label.x.npc = 0.2, color = "red")+
    labs(title = "Run006_IL2_Quanti")+
    theme_bw()
```

## validate correlation of predicted quant and software screenshot quant

```{r, eval=F}
colnames(test)
plt2pdt <- "MSC_Target48_Run008"
r2_pdt_quant <- sapply(analyts, function(x){
  if(x == "IL4"){
    predict(fit_list_48plx[[x]], newdata = list(NPX = f.list[[plt2pdt]]@assays@data$npx[rownames(f.list[[plt2pdt]]) == x, ]))
  }else{
    predict(fit_list_48plx[[x]], x = f.list[[plt2pdt]]@assays@data$npx[rownames(f.list[[plt2pdt]]) == x, ])$y
  }
})

f.list[[plt2pdt]]$Assay
dim(f.list[[plt2pdt]])

marker <- "IL10"
data.frame(Quant_screenshot = test$Run008_IL10_Quanti[-c(24, 30, 36, 42, 48)],
             Quant_predicted = 10^r2_pdt_quant[ , colnames(pdt_quant) == marker])%>%
  ggscatter(x = "Quant_screenshot", y = "Quant_predicted", palette = "jco",
   add = "reg.line", conf.int = F)+
   stat_cor(label.x.npc = 0.2, color = "red")+
    labs(title = "Run006_IL10_Quanti")+
    theme_bw()

```


